---
title: "BAYESIAN PROJECT"
author: "Ana Mendoza"
date: "2025-04-08"
output: word_document
---
load and clean the dataset 
```{r}
df <- read.csv("Autism Studies Dataset.csv")

df_clean <- df[!is.na(df$Sample.Size) & !is.na(df$Number.of.Cases), ]

df_clean$y <- df_clean$Number.of.Cases
df_clean$n <- df_clean$Sample.Size
```
prepare data for jags 
```{r}
data_list <- list(
  y = as.integer(df_clean$y),
  n = as.integer(df_clean$n),
  N = nrow(df_clean)
)
```
Here, we build the basic model where each study's observed cases follow a binomial distribution. We assume the true prevalence rate for each study is unknown, and we give it a flat (uniform) prior, meaning we don't assume anything ahead of time.
```{r}
library(rjags)


model_string <- "
model {
  for (i in 1:N) {
    y[i] ~ dbin(theta[i], n[i])
    theta[i] ~ dbeta(1, 1)
  }
}
"


jags_model <- jags.model(textConnection(model_string), data = data_list, n.chains = 3)
update(jags_model, 1000)  # Burn-in
samples <- coda.samples(jags_model, variable.names = c("theta"), n.iter = 5000)

```
This chunk plots the posterior distribution for just the first study. It shows what we believe the true prevalence rate is for that specific study after looking at the data. 
```{r}
posterior_matrix <- as.matrix(samples)

hist(posterior_matrix[, 1],
     main = "Posterior Distribution for Study 1",
     xlab = "Prevalence Rate",
     col = "lightblue", border = "white")
```


Posterior Summary for All Studies: 
Instead of looking at just one study, here we summarize the posterior results for all the studies by calculating the mean, standard deviation, and credible intervals for each estimated prevalence rate
```{r}
library(coda)

summary_stats <- summary(samples)
theta_stats <- summary_stats$statistics
theta_quantiles <- summary_stats$quantiles

posterior_summary <- data.frame(
  Mean = theta_stats[, "Mean"],
  SD = theta_stats[, "SD"],
  `2.5%` = theta_quantiles[, "2.5%"],
  `97.5%` = theta_quantiles[, "97.5%"]
)
head(posterior_summary)

```

 Plot Posterior Distributions for Multiple Studies: This chunk creates a graph showing the density curves for the first five studies’ posterior distributions. It lets us compare how different (or similar) the prevalence rates seem across studies.
```{r}

par(mfrow = c(2, 3))

for (i in 1:5) {
  
  max_value <- max(posterior_matrix[,i])
  
  hist(posterior_matrix[,i], 
       main = paste("Posterior for Study", i),
       xlab = "Prevalence Rate",
       col = "lightblue",
       border = "white",
       breaks = 20,
       xlim = c(0, max_value * 1.2))  # Zoom based on data
}


```
Estimate and Plot Overall Prevalence Rate (Pooled):Here we calculate the overall prevalence rate by pooling all the data together. We also plot it on top of the observed proportions to compare individual studies versus the pooled estimate.
```{r}
df_clean$p_hat <- df_clean$y / df_clean$n
pooled_p_hat <- sum(df_clean$y) / sum(df_clean$n)

hist(df_clean$p_hat, breaks=20, main="Observed Proportions vs. Pooled",
     xlab="Observed Prevalence", col="skyblue", border="white")
abline(v = pooled_p_hat, col="red", lwd=2, lty=2)
legend("topright", legend=paste("Pooled:", round(pooled_p_hat, 3)), col="red", lty=2)

```
Check Convergence Diagnostics: Before trusting our results, we use diagnostics to make sure the model actually converged and that the Markov chains are behaving nicely.
```{r}
gelman_diag <- gelman.diag(samples)
print(gelman_diag)

traceplot(samples[, 1:4])
```
Add a Hierarchical Beta-Binomial Model (Hyperprior):Instead of treating each study as completely independent, here we build a hierarchical model where all studies share a common Beta distribution, and we estimate its parameters (alpha and beta) too.

```{r}
model_hierarchical <- "
model {
  for (i in 1:N) {
    y[i] ~ dbin(theta[i], n[i])
    theta[i] ~ dbeta(alpha, beta)
  }
  
  alpha ~ dgamma(1, 0.01)
  beta ~ dgamma(1, 0.01)
}
"

jags_model_hier <- jags.model(textConnection(model_hierarchical), data = data_list, n.chains = 3)
update(jags_model_hier, 1000)
samples_hier <- coda.samples(jags_model_hier, variable.names = c("theta", "alpha", "beta"), n.iter = 5000)


```
Analyzing Hyperparameters and Posterior Densities:
After fitting the hierarchical model, this code looks at the estimated values of alpha and beta and plots the posterior distributions for a few studies under the new model.
```{r}
hier_matrix <- as.matrix(samples_hier)

alpha_post <- hier_matrix[, "alpha"]
beta_post <- hier_matrix[, "beta"]

cat("Posterior mean of alpha:", round(mean(alpha_post), 3), "\n")
cat("Posterior mean of beta:", round(mean(beta_post), 3), "\n")

# Plots
par(mfrow=c(1,2))
hist(alpha_post, main="Posterior of alpha", col="lightcoral", xlab="alpha", border="white")
hist(beta_post, main="Posterior of beta", col="lightblue", xlab="beta", border="white")

par(mfrow=c(1,1))
plot(density(hier_matrix[, "theta[1]"]), main="Posterior Densities (Hierarchical Model)",
     xlab="Prevalence Rate", ylim=c(0, 15), col=1)
for (i in 2:5) {
  lines(density(hier_matrix[, paste0("theta[", i, "]")]), col=i)
}
legend("topright", legend=paste("Study", 1:5), col=1:5, lty=1)

```
Posterior Predictive Checks (PPC):
Posterior predictive checks are like a "reality check" , we simulate new fake datasets from our model and see if they match the real-world data we observed.
```{r}
model_ppc <- "
model {
  for (i in 1:N) {
    y[i] ~ dbin(theta[i], n[i])
    y_rep[i] ~ dbin(theta[i], n[i])
    theta[i] ~ dbeta(1, 1)
  }
}
"

data_list_ppc <- data_list
jags_ppc <- jags.model(textConnection(model_ppc), data = data_list_ppc, n.chains = 3)
update(jags_ppc, 1000)
ppc_samples <- coda.samples(jags_ppc, variable.names = c("y_rep"), n.iter = 5000)

y_rep_matrix <- as.matrix(ppc_samples)
y_rep_means <- colMeans(y_rep_matrix)

plot(df_clean$y, y_rep_means, main="Posterior Predictive Check",
     xlab="Observed Cases", ylab="Predicted Cases", col="blue", pch=16)
abline(0, 1, col="red", lty=2)

```
Mapping Autism Prevalence by Country:
If our dataset includes country names, we can plot the estimated prevalence rates across the world to spot geographic patterns.
```{r}
library(ggplot2)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)

country_means <- df_clean %>%
  mutate(theta_mean = colMeans(posterior_matrix)) %>%
  group_by(Country) %>%
  summarise(MeanPrevalence = mean(theta_mean))

world <- ne_countries(scale = "medium", returnclass = "sf")

world_map <- left_join(world, country_means, by = c("name" = "Country"))

ggplot(data = world_map) +
  geom_sf(aes(fill = MeanPrevalence)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(title = "Estimated Autism Prevalence by Country",
       fill = "Prevalence") +
  theme_minimal()



```
